Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-05-06 writeups/Su_IGW_break.tex
check in accepted version

\date{Accepted XXX\@. Received YYY\@; in original form ZZZ}

@@

\begin{abstract}
    In binaries composed of \textcolor{Corr}{either early-type stars or white
    dwarfs}, the dominant tidal process involves the excitation of internal
    gravity waves (IGWs), which propagate towards the stellar surface, and their
    dissipation via nonlinear wave breaking. We perform 2D hydrodynamical
    simulations of this wave breaking process in a stratified, isothermal
    atmosphere. We find that, after an initial transient phase, the dissipation
    of the IGWs naturally generates a sharp critical layer, separating the lower
    stationary region (with no mean flow) and the upper ``synchronized'' region
    (with the mean flow velocity equal to the horizontal wave phase speed).
    While the critical layer is steepened by absorption of these waves, it is
    simultaneously broadened by Kelvin-Helmholtz instabilities such that, in
    steady state, the critical layer width is determined by the Richardson
    criterion. We study the absorption and reflection of incident waves off the
    critical layer and provide analytical formulae describing its long-term
    evolution. The result of this study is important for characterizing the
    evolution of tidally heated white dwarfs and other binary stars.
\end{abstract}

@@

We consider a incompressible, isothermally stratified fluid representing a
stellar envelope or atmosphere. We study dynamics in 2D, so that fluid variables
depend only on the Cartesian coordinates $x$ and $z$. \textcolor{Corr}{While it
is well known that waves break differently in 2D versus 3D} \citep{klostermeyer,
winters1994}, the dynamical effect of the breaking process is likely to be
similar in 2D \citep{barker_ogilvie}. We approximate the gravitational field as
uniform, pointing in the $(-\uv{z})$ direction. The plane-parallel approximation
is justified since wave breaking generally occurs near the stellar surface. The
background density stratification is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density. We denote background quantities
with overbars and perturbation quantities with primes.

The Euler equations for an incompressible fluid in a uniform gravitational field
are
% \footnote{\textcolor{Corr}{These equations arise by enforcing fluid
% incompressibility ($\mdil{\rho}{t} = 0$ or the adiabatic index $\gamma \to
% \infty$) in the Euler equations. They also can be derived from the commonly used
% ``sound-proof'' \emph{anelastic equations} \citep[see e.g.][]{ogura1962scale,
% sutherland1} in the limit where (i) the density stratification is weak ($H \gg
% L$ for perturbation wavelength $L$) and (ii) the fluid is incompressible (note
% the anelastic equations are defined for arbitrary $\gamma$). Under condition
% (i), the anelastic approximation reduces to the Boussinesq approximation
% \citep{klein2009asymptotics}. Our equations permit a weak but nonzero density
% stratification and can be interpreted as ``locally Boussinesq''. Condition (ii)
% filters out sound waves. The dynamical effect of wave breaking is likely to be
% similar for different sets of fluid equations.}}
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $(-g\uv{z})$. \textcolor{Corr}{Note that these equations
conserve the same wave energy as the commonly used anelastic equations
\citep{ogura1962scale,anel_part1} and thus give the same wave amplitude
growth. Appendix~\ref{s:equation_deriv} provides a derivation of these
equations and justification for using them.}

For this isothermal background, hydrostatic equilibrium implies
\textcolor{Corr}{$\overline{P}(z) = \overline{\rho}(z) g H$}. We assume there is
initially no background flow, so $\bm{u} = \bm{u}'$. Physically, this assumption
corresponds to a non-rotating star.

@@

In the small perturbation limit, we may linearize Eq.~\eqref{se:nl_upsilon}.
\textcolor{Corr}{The resulting equations admit the canonical IGW solution}
\citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency}, $N$, is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
and is constant. Other dynamical quantities are simply related to $u'_z$.

@@

    \item The phase and group velocities are given by:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \p{2H}^{-2}},\\
            \bm{c}_{g} &= N\frac{\s{k_{z}^2 + \p{2H}^{-2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\s{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}^{3/2}}.\label{eq:vg}
        \end{align}
        \textcolor{Corr}{The additional $\p{2H}^{-2}$ term in the denominator
        accounts for the growing amplitude of the IGW in the $z$ direction (as
        the $z$ wavenumber is effectively $k_z - i / (2H)$).} We note
        $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\s{\p{k_{z}H}^{-2}} \approx 0$.
        In the Boussinesq approximation where terms of order
        $\mathcal{O}\p{H^{-2}}$ are ignored, the phase and group velocities are
        exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

@@

Consider now the long-term evolution of the critical layer due to continuous
horizontal momentum transfer by IGWs. \textcolor{Corr}{Any incident horizontal
momentum flux absorbed by the fluid, denoted $F_a(t)$, must manifest as
additional horizontal momentum of the shear flow}.
\textcolor{Corr}{Additionally, as the mean flow $\overline{U}$ cannot grow
efficiently above $\overline{U}_c$ (due to the breakdown of the linear
solution), we assume $\overline{U}$ saturates at $\overline{U}_c$, which holds
to good accuracy (see Fig.~\ref{fig:nl_fluxes}). In this case, the critical
layer must propagate downward in response to the incident momentum flux.} The
horizontal momentum of the shear flow satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - F_a(t) = 0.
\end{equation}
Assuming $\overline{U}(z > z_c) \approx \overline{U}_c$ and $\overline{U}(z <
z_c) \approx 0$, this condition becomes
\begin{equation}
    -\overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = F_a(t).\label{eq:zc_anal}
\end{equation}
If $F_a$ is constant in time, the height of the critical layer $z_c(t)$ has
analytical solution:
\begin{equation}
    z_c(t) = -H\ln \s{\exp\p{-\frac{z_c(t = 0)}{H}} +
        \frac{tF_a}{\overline{U}_c H\overline{\rho}_0}},\label{eq:zc_sol}
\end{equation}
where $z_c(t = 0)$ is the initial critical layer height.

@@

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from \textcolor{Corr}{the minimum of $0.1 / N$
and the advective Courant-Friedrichs-Lewy (CFL) time. The CFL time is given by}
$\Delta t = 0.7 \min(\Delta x / u_x,\Delta z / u_{z})$, where the minimum is
taken over every grid point in the domain, and $\Delta x \equiv L_x / N_x$ and
$\Delta z \equiv L_z / N_z$ are the grid spacings in the $x$ and $z$ directions
respectively.

@@

We aim to damp disturbances that reach the vertical boundaries of the simulation
domain without inducing nonphysical reflection. To do so, we replace material
derivatives in Eq.~\eqref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.3H$ and $z_T = 9.5H$ are the boundaries of the lower and upper
damping zones respectively. This damps perturbations below $z_B$ and
above $z_T$ with damping time $\tau$ and negligibly affects the dynamics between
$z_B$ and $z_T$. \textcolor{Corr}{Most importantly, horizontal momentum remains
conserved between $z_B$ and $z_T$, and outgoing boundary conditions are imposed
at $z_B, z_T$}. We choose the transition width $\Delta z = 0.25H$ and damping
time $\tau = 1 / (15N)$. This prescription is similar to \citet{lecoanet_damp}
and has the advantage of being smooth, important for spectral methods. Further
details of our implementation of the fluid equations in Dedalus are described in
Appendix~\ref{se:strat_impl}.

@@

We expect the waves to follow the analytical solution given by
Eq.~\eqref{eq:uz_lin} and the corresponding $u_x(x, z, t)$; we denote this
analytical solution $\bm{u}_{an}(x, z, t)$. The amplitude of the observed IGW in the
simulation field $\bm{u}$ relative to $\bm{u}_{an}$ over
some region $z \in [z_b, z_t]$ can be estimated from
\begin{equation}
    A_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{an}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{an}}^2\;\mathrm{d}x\mathrm{d}z}.
        \label{eq:ahat_def}
\end{equation}
The subscript $i$ denotes the incident wave. If $\bm{u} = \bm{u}_{an}$, then
$A_i(t) = 1$. \textcolor{Corr}{The factor of $\bar{\rho}$ inside the integrands
in Eq.~\eqref{eq:ahat_def} corrects for the $\propto e^{z/2H}$ growth of
$\bm{u}_{an}$; without it, $A_i(t)$ would be dominated by the contribution near
$z_t$.}

For the weakly forced simulation, we expect
$A_i(t) = 1$ when integrated between the forcing and damping zones, i.e.\
$z_b \gtrsim z_0$ and $z_t \lesssim z_T$ ($z_0, z_T$ are defined in
Eq.~\eqref{eq:vol_drive} and Eq.~\eqref{eq:Gamma} respectively). For consistency
with the nonlinear case later, we choose $z_b = z_0 + 3\sigma$ and $z_t = z_b +
H$. Note that using a larger integration domain by choosing $z_t = z_T - \Delta
z$ just below the upper damping zone instead does not change the measured $A_i$.
The resulting measurement of $A_i(t)$ is shown in Fig.~\ref{fig:lin_amps}, and
indeed $A_i \approx 1$ after the initial transient.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of the excited IGW over time (in units of $N^{-1}$) in
    the weakly forced simulation, computed using Eq.~\eqref{eq:ahat_def}.
    $A_i(t) = 1$ corresponds to perfect agreement with the analytical estimate.
    After an initial transient phase, we observe $A_i(t)$ asymptotes to $\approx
    1$, implying continuous excitation of identical IGW with the expected
    amplitude. The small deviation of $A_i(t)$ from unity may be due to
    \textcolor{Corr}{truncation error in our implicit timestepping scheme}, as a
    relatively large fixed step size $\Delta t = 0.1/N$ was used for this
    simulation.}\label{fig:lin_amps}
\end{figure}

@@

To perform simulations of wave breaking phenomena, we use the same setup as
described in Section~\ref{s:numerics} and Section~\ref{s:weak_sim} except for
different values of $C$ and $\nu$. In particular, we choose $C$ such that
$\abs{\xi_z k_z} = 0.1$ in the forcing zone ($z = z_0$). The linear solution
predicts $\abs{\xi_z k_z} \sim 4.25$ at the upper damping zone $z_T$.
\textcolor{Corr}{We choose the viscosity $\nu$ for each resolution to be as
small as possible while still resolving the shortest spatial scales of the wave
breaking.} A table of our simulations can be found in Table~\ref{tab:params}.
\begin{table}
    \centering
    \begin{tabular}{l c c c}
        Resolution & $\mathrm{Re}$\\\bottomrule
        $1024 \times 4096$ & $2048$\\
        $768 \times 3072$ & $1024$\\
        $512 \times 2048$ & $512$\\
        $256 \times 1024$ & $341$\\
        $256 \times 1024$ & $205$\\
        $256 \times 1024$ & $146$\\
    \end{tabular}
    \caption{Spectral resolutions and Reynolds numbers of simulations of wave
    breaking.}\label{tab:params}
\end{table}

@@

% convert yubo_000054.png -crop 2000x2000+150+250 out.png
In Fig.~\ref{fig:snapshots}, we present snapshots of $u_x$ and $\Upsilon$ at
various phases of the simulation. \textcolor{Corr}{Note that $\Upsilon \lesssim
0.1$, so the density stratification does not deviate significantly from
equilibrium.} The flow evolves through several distinct stages:
\begin{enumerate}
    \item At early times (top left panel), the flow resembles a linear IGW lower
        in the simulation domain but breaks down into smaller-scale features at
        higher $z$. Some characteristic swirling motion can be seen in the
        advected scalar $\Upsilon$, indicating Kelvin-Helmholtz instabilities.

@@

    \item At the end of the simulation (bottom right panel), the
        \textcolor{Corr}{critical layer has advanced downwards, but otherwise}
        the flow shows very few significant qualitative differences from the
        previous snapshot. This suggests that the latter phase of the simulation
        has reached a steady state. \textcolor{Corr}{Notably, the horizontal
        banded structure of $u_x$ in the upper, synchronized fluid does not
        continue to evolve (also visible in the top panel of
        Fig.~\ref{fig:nl_fluxes}), suggesting that momentum redistribution and
        mixing within the synchronized fluid are negligible.}
\end{enumerate}

@@

    \includegraphics[width=0.45\textwidth]{plots/yubo_000134_labeled.png}\hfil
    %
    \includegraphics[width=0.45\textwidth]{plots/yubo_000451_labeled.png}
    \caption{Snapshots of $u_x$ and $\Upsilon \equiv \ln \p{\rho / \bar{\rho}}$
    in the fiducial simulation illustrating distinct phases of the evolution of
    the flow. \textcolor{Corr}{See the online PDF for a color version.} Damping
    layers \textcolor{Corr}{at the top and bottom of the simulation domain} are
    shaded in light grey (see Section~\ref{ss:damping}), while the forcing zone
    \textcolor{Corr}{in the lower middle portion of the simulation domain} (see
    Section~\ref{ss:forcing}) is shaded in light green (boundaries are at $z_0
    \pm 3\sigma$). The four panels illustrate (i) the initial transient wave
    breaking phase, (ii) formation of a distinct critical layer, (iii)
    steepening of the critical layer, and (iv) downward advance of the critical
    layer.}\label{fig:snapshots}
\end{figure*}

@@

The formation of the critical layer is associated with a strong shear flow. What
is the width of this layer? Inspection of Fig.~\ref{fig:snapshots} suggests the
\textcolor{Corr}{presence} of the Kelvin-Helmholtz Instability (KHI) in the
critical layer. In a stratified medium, KHI occurs when the Richardson number
(Eq.~\eqref{eq:ri_def}) satisfies $\mathrm{Ri} \lesssim 1/4$
\citep[e.g.][]{shu1991physics}. It is natural to suspect that the shear flow
cannot steepen further than the onset of KHI\@. To test this, we compute the
local $\mathrm{Ri}$ for the shear flow around the critical layer.

>>>>

\date{Accepted 2020 May 06. Received 2020 April 13; in original form 2020
February 26}

@@

\begin{abstract}
    In binaries composed of either early-type stars or white
    dwarfs, the dominant tidal process involves the excitation of internal
    gravity waves (IGWs), which propagate towards the stellar surface, and their
    dissipation via nonlinear wave breaking. We perform 2D hydrodynamical
    simulations of this wave breaking process in a stratified, isothermal
    atmosphere. We find that, after an initial transient phase, the dissipation
    of the IGWs naturally generates a sharp critical layer, separating the lower
    stationary region (with no mean flow) and the upper ``synchronized'' region
    (with the mean flow velocity equal to the horizontal wave phase speed).
    While the critical layer is steepened by absorption of these waves, it is
    simultaneously broadened by Kelvin-Helmholtz instabilities such that, in
    steady state, the critical layer width is determined by the Richardson
    criterion. We study the absorption and reflection of incident waves off the
    critical layer and provide analytical formulae describing its long-term
    evolution. The result of this study is important for characterizing the
    evolution of tidally heated white dwarfs and other binary stars.
\end{abstract}

@@

We consider a incompressible, isothermally stratified fluid representing a
stellar envelope or atmosphere. We study dynamics in 2D, so that fluid variables
depend only on the Cartesian coordinates $x$ and $z$. While it
is well known that waves break differently in 2D versus 3D \citep{klostermeyer,
winters1994}, the dynamical effect of the breaking process is likely to be
similar in 2D \citep{barker_ogilvie}. We approximate the gravitational field as
uniform, pointing in the $(-\uv{z})$ direction. The plane-parallel approximation
is justified since wave breaking generally occurs near the stellar surface. The
background density stratification is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density. We denote background quantities
with overbars and perturbation quantities with primes.

The Euler equations for an incompressible fluid in a uniform gravitational field
are
% \footnote{These equations arise by enforcing fluid
% incompressibility ($\mdil{\rho}{t} = 0$ or the adiabatic index $\gamma \to
% \infty$) in the Euler equations. They also can be derived from the commonly used
% ``sound-proof'' \emph{anelastic equations} \citep[see e.g.][]{ogura1962scale,
% sutherland1} in the limit where (i) the density stratification is weak ($H \gg
% L$ for perturbation wavelength $L$) and (ii) the fluid is incompressible (note
% the anelastic equations are defined for arbitrary $\gamma$). Under condition
% (i), the anelastic approximation reduces to the Boussinesq approximation
% \citep{klein2009asymptotics}. Our equations permit a weak but nonzero density
% stratification and can be interpreted as ``locally Boussinesq''. Condition (ii)
% filters out sound waves. The dynamical effect of wave breaking is likely to be
% similar for different sets of fluid equations.}
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $(-g\uv{z})$. Note that these equations
conserve the same wave energy as the commonly used anelastic equations
\citep{ogura1962scale,anel_part1} and thus give the same wave amplitude
growth. Appendix~\ref{s:equation_deriv} provides a derivation of these
equations and justification for using them.

For this isothermal background, hydrostatic equilibrium implies
$\overline{P}(z) = \overline{\rho}(z) g H$. We assume there is
initially no background flow, so $\bm{u} = \bm{u}'$. Physically, this assumption
corresponds to a non-rotating star.

@@

In the small perturbation limit, we may linearize Eq.~\eqref{se:nl_upsilon}.
The resulting equations admit the canonical IGW solution
\citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency}, $N$, is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
and is constant. Other dynamical quantities are simply related to $u'_z$.

@@

    \item The phase and group velocities are given by:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \p{2H}^{-2}},\\
            \bm{c}_{g} &= N\frac{\s{k_{z}^2 + \p{2H}^{-2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\s{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}^{3/2}}.\label{eq:vg}
        \end{align}
        The additional $\p{2H}^{-2}$ term in the denominator
        accounts for the growing amplitude of the IGW in the $z$ direction (as
        the $z$ wavenumber is effectively $k_z - i / (2H)$). We note
        $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\s{\p{k_{z}H}^{-2}} \approx 0$.
        In the Boussinesq approximation where terms of order
        $\mathcal{O}\p{H^{-2}}$ are ignored, the phase and group velocities are
        exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

@@

Consider now the long-term evolution of the critical layer due to continuous
horizontal momentum transfer by IGWs. Any incident horizontal
momentum flux absorbed by the fluid, denoted $F_a(t)$, must manifest as
additional horizontal momentum of the shear flow.
Additionally, as the mean flow $\overline{U}$ cannot grow
efficiently above $\overline{U}_c$ (due to the breakdown of the linear
solution), we assume $\overline{U}$ saturates at $\overline{U}_c$, which holds
to good accuracy (see Fig.~\ref{fig:nl_fluxes}). In this case, the critical
layer must propagate downward in response to the incident momentum flux. The
horizontal momentum of the shear flow satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - F_a(t) = 0.
\end{equation}
Assuming $\overline{U}(z > z_c) \approx \overline{U}_c$ and $\overline{U}(z <
z_c) \approx 0$, this condition becomes
\begin{equation}
    -\overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = F_a(t).\label{eq:zc_anal}
\end{equation}
If $F_a$ is constant in time, the height of the critical layer $z_c(t)$ has
analytical solution:
\begin{equation}
    z_c(t) = -H\ln \s{\exp\p{-\frac{z_c(t = 0)}{H}} +
        \frac{tF_a}{\overline{U}_c H\overline{\rho}_0}},\label{eq:zc_sol}
\end{equation}
where $z_c(t = 0)$ is the initial critical layer height.

@@

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from the minimum of $0.1 / N$
and the advective Courant-Friedrichs-Lewy (CFL) time. The CFL time is given by
$\Delta t = 0.7 \min(\Delta x / u_x,\Delta z / u_{z})$, where the minimum is
taken over every grid point in the domain, and $\Delta x \equiv L_x / N_x$ and
$\Delta z \equiv L_z / N_z$ are the grid spacings in the $x$ and $z$ directions
respectively.

@@

We aim to damp disturbances that reach the vertical boundaries of the simulation
domain without inducing nonphysical reflection. To do so, we replace material
derivatives in Eq.~\eqref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.3H$ and $z_T = 9.5H$ are the boundaries of the lower and upper
damping zones respectively. This damps perturbations below $z_B$ and
above $z_T$ with damping time $\tau$ and negligibly affects the dynamics between
$z_B$ and $z_T$. Most importantly, horizontal momentum remains
conserved between $z_B$ and $z_T$, and outgoing boundary conditions are imposed
at $z_B, z_T$. We choose the transition width $\Delta z = 0.25H$ and damping
time $\tau = 1 / (15N)$. This prescription is similar to \citet{lecoanet_damp}
and has the advantage of being smooth, important for spectral methods. Further
details of our implementation of the fluid equations in Dedalus are described in
Appendix~\ref{se:strat_impl}.

@@

We expect the waves to follow the analytical solution given by
Eq.~\eqref{eq:uz_lin} and the corresponding $u_x(x, z, t)$; we denote this
analytical solution $\bm{u}_{an}(x, z, t)$. The amplitude of the observed IGW in the
simulation field $\bm{u}$ relative to $\bm{u}_{an}$ over
some region $z \in [z_b, z_t]$ can be estimated from
\begin{equation}
    A_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{an}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{an}}^2\;\mathrm{d}x\mathrm{d}z}.
        \label{eq:ahat_def}
\end{equation}
The subscript $i$ denotes the incident wave. If $\bm{u} = \bm{u}_{an}$, then
$A_i(t) = 1$. The factor of $\bar{\rho}$ inside the integrands
in Eq.~\eqref{eq:ahat_def} corrects for the $\propto e^{z/2H}$ growth of
$\bm{u}_{an}$; without it, $A_i(t)$ would be dominated by the contribution near
$z_t$.

For the weakly forced simulation, we expect
$A_i(t) = 1$ when integrated between the forcing and damping zones, i.e.\
$z_b \gtrsim z_0$ and $z_t \lesssim z_T$ ($z_0, z_T$ are defined in
Eq.~\eqref{eq:vol_drive} and Eq.~\eqref{eq:Gamma} respectively). For consistency
with the nonlinear case later, we choose $z_b = z_0 + 3\sigma$ and $z_t = z_b +
H$. Note that using a larger integration domain by choosing $z_t = z_T - \Delta
z$ just below the upper damping zone instead does not change the measured $A_i$.
The resulting measurement of $A_i(t)$ is shown in Fig.~\ref{fig:lin_amps}, and
indeed $A_i \approx 1$ after the initial transient.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of the excited IGW over time (in units of $N^{-1}$) in
    the weakly forced simulation, computed using Eq.~\eqref{eq:ahat_def}.
    $A_i(t) = 1$ corresponds to perfect agreement with the analytical estimate.
    After an initial transient phase, we observe $A_i(t)$ asymptotes to $\approx
    1$, implying continuous excitation of identical IGW with the expected
    amplitude. The small deviation of $A_i(t)$ from unity may be due to
    truncation error in our implicit timestepping scheme, as a
    relatively large fixed step size $\Delta t = 0.1/N$ was used for this
    simulation.}\label{fig:lin_amps}
\end{figure}

@@

To perform simulations of wave breaking phenomena, we use the same setup as
described in Section~\ref{s:numerics} and Section~\ref{s:weak_sim} except for
different values of $C$ and $\nu$. In particular, we choose $C$ such that
$\abs{\xi_z k_z} = 0.1$ in the forcing zone ($z = z_0$). The linear solution
predicts $\abs{\xi_z k_z} \sim 4.25$ at the upper damping zone $z_T$.
We choose the viscosity $\nu$ for each resolution to be as
small as possible while still resolving the shortest spatial scales of the wave
breaking. A table of our simulations can be found in Table~\ref{tab:params}.
\begin{table}
    \centering
    \begin{tabular}{l c c c}
        Resolution & $\mathrm{Re}$\\\bottomrule
        $1024 \times 4096$ & $2048$\\
        $768 \times 3072$ & $1024$\\
        $512 \times 2048$ & $512$\\
        $256 \times 1024$ & $341$\\
        $256 \times 1024$ & $205$\\
        $256 \times 1024$ & $146$\\
    \end{tabular}
    \caption{Spectral resolutions and Reynolds numbers of simulations of wave
    breaking.}\label{tab:params}
\end{table}

@@

% convert yubo_000054.png -crop 2000x2000+150+250 out.png
In Fig.~\ref{fig:snapshots}, we present snapshots of $u_x$ and $\Upsilon$ at
various phases of the simulation. Note that $\Upsilon \lesssim
0.1$, so the density stratification does not deviate significantly from
equilibrium. The flow evolves through several distinct stages:
\begin{enumerate}
    \item At early times (top left panel), the flow resembles a linear IGW lower
        in the simulation domain but breaks down into smaller-scale features at
        higher $z$. Some characteristic swirling motion can be seen in the
        advected scalar $\Upsilon$, indicating Kelvin-Helmholtz instabilities.

@@

    \item At the end of the simulation (bottom right panel), the
        critical layer has advanced downwards, but otherwise
        the flow shows very few significant qualitative differences from the
        previous snapshot. This suggests that the latter phase of the simulation
        has reached a steady state. Notably, the horizontal
        banded structure of $u_x$ in the upper, synchronized fluid does not
        continue to evolve (also visible in the top panel of
        Fig.~\ref{fig:nl_fluxes}), suggesting that momentum redistribution and
        mixing within the synchronized fluid are negligible.
\end{enumerate}

@@

    \includegraphics[width=0.45\textwidth]{plots/yubo_000134_labeled.png}\hfil
    %
    \includegraphics[width=0.45\textwidth]{plots/yubo_000451_labeled.png}
    \caption{Snapshots of $u_x$ and $\Upsilon \equiv \ln \p{\rho / \bar{\rho}}$
    in the fiducial simulation illustrating distinct phases of the evolution of
    the flow. See the online PDF for a color version. Damping
    layers at the top and bottom of the simulation domain are
    shaded in light grey (see Section~\ref{ss:damping}), while the forcing zone
    in the lower middle portion of the simulation domain (see
    Section~\ref{ss:forcing}) is shaded in light green (boundaries are at $z_0
    \pm 3\sigma$). The four panels illustrate (i) the initial transient wave
    breaking phase, (ii) formation of a distinct critical layer, (iii)
    steepening of the critical layer, and (iv) downward advance of the critical
    layer.}\label{fig:snapshots}
\end{figure*}

@@

The formation of the critical layer is associated with a strong shear flow. What
is the width of this layer? Inspection of Fig.~\ref{fig:snapshots} suggests the
presence of the Kelvin-Helmholtz Instability (KHI) in the
critical layer. In a stratified medium, KHI occurs when the Richardson number
(Eq.~\eqref{eq:ri_def}) satisfies $\mathrm{Ri} \lesssim 1/4$
\citep[e.g.][]{shu1991physics}. It is natural to suspect that the shear flow
cannot steepen further than the onset of KHI\@. To test this, we compute the
local $\mathrm{Ri}$ for the shear flow around the critical layer.

It is difficult to accurately measure the Richardson number, as it depends on
the derivative of the velocity. We measure $\mathrm{Ri}$ as follows: we first
assign an $\mathrm{Ri}_x(x, t)$ for every $x$ in the critical layer, then take
the median of $\mathrm{Ri}$ for the entire layer. $\mathrm{Ri}_x$ is computed
using the vertical distance over which the local $u_x$ increases from
$0.3\bar{U}_c$ to $\bar{U}_c$ (see Eq.~\eqref{eq:u_crit}). The value $0.3$ is
necessary to exclude the small mean flow generated in the weakly nonlinear
regime far below the
critical layer. This procedure can be written:
\begin{align}
    z_{CL, \min}(x, t) &\equiv \min
        \z{z\mid u_x(x, z, t) > 0.3\overline{U}_c},\\
    z_{CL, \max}(x, t) &\equiv \max
        \z{z\mid u_x(x, z, t) < \overline{U}_c},\\
    \mathrm{Ri}_x(x, t) &\equiv
        \p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}},\\
    \mathrm{Ri}(t) &\equiv \med_x\mathrm{Ri}\p{x, t}.\label{eq:ri_med_def}
\end{align}
We use the background buoyancy frequency to compute $\mathrm{Ri}$, as
fluctuations do not change $N^2$ significantly ($\sim 1\%$). To understand the
variation in $\mathrm{Ri}$ over $x$, we also compute $\min\limits_x
\mathrm{Ri}_x(x, t)$ (the maximum is very noisy). Both are shown in
Fig.~\ref{fig:nl_f_ri}. Absorption of incident IGWs quickly
decreases the Richardson number to between $0.25$ and $0.5$, characteristic of
the onset of the KHI\@.
